Supplementary Materials for \"COATi: statistical pairwise alignment of protein coding sequences\

### 2024-04-26 alignpair_letter_source.tex
Mark revision 2

% !TEX program=lualatex
\documentclass[12pt,letterpaper]{article}

@@

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage[warnings-off={mathtools-colon,mathtools-overbracket}]{unicode-math}
\usepackage{fontspec}
\usepackage{microtype}

\setmainfont{TeX Gyre Pagella}
\setmathfont{Latin Modern Math}

@@

% document begins here
\begin{document}
% title goes here:
\begin{flushleft}
{\Large\textbf{COATi: statistical pairwise alignment of protein-coding sequences}}
\newline
% authors go here:
\\
Juan J. García Mesa\textsuperscript{1,2},
Ziqi Zhu\textsuperscript{1,3},
Reed A. Cartwright\textsuperscript{1,3,*}\\
\bigskip
1 The Biodesign Institute, Arizona State University, Tempe, Arizona, USA\\
2 Ira A.\ Fulton Schools of Engineering, Arizona State University, Tempe, Arizona, USA\\
3 School of Life Sciences, Arizona State University, Tempe, Arizona, USA\\
\bigskip
* cartwright@asu.edu

@@

\begin{abstract}
\noindent
Sequence alignment is an essential method in bioinformatics and the basis of many analyses, including phylogenetic inference, ancestral sequence reconstruction, and gene annotation. Sequence artifacts and errors made during alignment reconstruction can impact downstream analyses leading to erroneous conclusions in comparative and functional genomic studies. For example, abiological frameshifts and early stop codons are common artifacts found in protein coding sequences that have been annotated in reference genomes. More significantly, while indels can occur both within and between codons in natural sequences, most amino-acid- and codon-based aligners assume that indels only occur between codons. This mismatch between biology and alignment algorithms produces sub-optimal alignments and errors in downstream analyses.

To address these issues, we present COATi, a statistical, codon-aware pairwise aligner that supports complex insertion-deletion models and can handle artifacts present in genomic data. COATi allows users to reduce the amount of discarded data while generating more accurate sequence alignments. COATi can infer indels both within and between codons, leading to improved sequences alignments. We applied COATi to a dataset containing orthologous protein-coding sequences from humans and gorillas, and conclude that 41\% of indels occurred between codons, agreeing with previous work in rodents. We also applied to semi-empirical benchmark alignments and find that COATi outperforms several popular alignment programs on several measures of alignment quality and accuracy.

@@

Within protein-coding sequences, indels may occur in between any pair of adjacent nucleotides, and therefore, gaps in alignments of natural sequences may occur both between and within codons (Fig.\ \ref{fig:aln}). Gaps that occur after the first, second, and third positions of a codon are known as phase-1, phase-2, and phase-3 gaps, respectively. Notably, phase-3 gaps occur between codons and are also known as phase-0 gaps \citep[e.g.][]{taylor2004occurrence}. Indels that occur within codons can produce an amino-acid change in addition to adding or removing nucleotides (Fig.\ \ref{fig:aln}), and due to the structure of the genetic code, phase-1 gaps are less common than phase-2 which are less common than phase-3 \citep{taylor2004occurrence, zhu2022profiling}. While all three phases of gaps occur in natural sequences, alignments performed in amino-acid or codon space force all gaps to be phase-3 gaps. Because only about 42\% of indels are phase-3 \citep{taylor2004occurrence, zhu2022profiling}, this mismatch between aligner assumptions and biology can produce sub-optimal alignments and inflated estimates of sequence divergence
\citep[Fig.\ \ref{fig:aln};][]{redelings2007incorporating}.

% Figures are 86 mm or 178 mm wide
\begin{figure}[h!]
    \centering%
    \includegraphics[scale=1]{figures/fig-aln.pdf}
    \par
    \caption{
        Standard algorithms produce suboptimal alignments.
        (a) shows the true alignment of an ancestor sequence (A) and a descendant sequence (D).
        (b--d) are the results of different aligners. Nucleotide mismatches are highlighted in red.---Notably, COATi is the only aligner able to retrieve the biological alignment in this example.---%
        %
        Indels in protein-coding sequences can be classified as having one of three difference phases and being one of two different types.
        Phases refer the the location of the gap with respect to the reading frame, while types refer to the consequence of the indel.
        Phase-1, phase-2, and phase-3 indels are shown in blue, orange, and green respectively.
        Additionally, the orange indel is type-II (an amino-acid indel plus an amino-acid change) while the blue indel is type-I (an amino-acid indel only). The difference between an in-frame and a frameshift indel is not displayed.
        }
    \label{fig:aln}
\end{figure}

@@

\begin{figure}[!ht]
    \centering
    \includegraphics[width=\textwidth]{figures/fig-fst-base-calling-new.pdf}
    \caption[FSTs more the generation of sequences]{%
    Finite state transducers (FSTs) model the generation of an output sequence based on an input sequence.
    %
    (a) A graph of a probabilistic FST \citep{cotterell-etal-2014-stochastic} for base-calling errors using a Mealy-machine architecture, where parameter $u$ is the error rate. This graph contains two states (S and M) connected by arcs, with labels ``input symbols~:~output symbols~/~weight''. Arcs consume symbols from the input sequence and emit symbols to the output sequence. Weights describe the probability that an arc is taken given the input symbols. Epsilon (ε) is a special symbol denoting that no symbols were either consumed or emitted.
    %
    (b) An FST for matching sequences against ambiguous nucleotides (N). 
    (c) An FST that results from the composition ($∘$ operation) of the Error FST with the Ambiguity FST.
    %
    %Let $x$ be a symbol in the input of (a), $y$ be a symbol in the output of (b), and $z$ be a symbol in both the output of (a) and the input of (b). An arc in the composed FST (c) has an input symbol $x$, an output symbol $y$, and a weight that is the combination over $z$ of the aggregate weights of all $x$-$z$-$y$ arc pairs.
    %
    }\label{fig:base-calling}
\end{figure}

@@

COATi FST is not a true probabilistic FST and cannot be used as-is to simulate output sequences based on an input sequence. This is because it is missing a parameter to control how often Ns are added to the output sequence. This is a design feature to allow COATi FST to properly weight ambiguous nucleotides as representing any other symbol. In addition, COATi's indel FST (Fig.~\ref{fig:coati-fst}b) implicitly assumes that match states occur both immediately before and immediately after the alignment. This allows COATi FST to assign the same weight to gaps at the beginning and end of an alignment but also introduces a normalization constant (not shown) to reflect that mass is lost by not allowing the indel FST to terminate from node C. While this normalization constant is not needed to find the most likely alignment or to sample alignments, it can be calculated using a Markov model that considers only the transitions between states M and D in the indel FST (Fig.~\ref{fig:coati-fst}b) and ignoring any states in between:
\[
\frac{1-e + g(e-g)^{n+1}}{1-e+g} \approx \frac{1-e}{1-e+g} 
\]
where $g$ is the gap opening parameter, $e$ is the gap-extension parameter, and $n$ is the length of the input sequence. The exponent has a $+1$ due to the presence of an ``immortal match'' immediately after the input sequence.

@@

\begin{figure}[h!]
\includegraphics[width=\textwidth]{figures/fig-fst-coati.pdf}
\caption{The COATi FST is built from simpler FSTs via composition.
(a) The substitution FST encodes a $61 \times 61 $ codon substitution model with 3721 arcs from S to M. These arcs consume three nucleotides from the input tape and emit three nucleotides to the output tape. The weight of each arc is a conditional probability derived from a codon substitution model. See Fig.~\ref{fig:base-calling} for more details about reading this graph.
%
(b) The indel FST allows for insertions (H to I) and deletions (C to D). Here $g$ is the gap-opening parameter and $e$ is the gap-extension parameter.
Insertion arcs are weighted according to the codon model's stationary distribution of nucleotides, and deletion arcs have a weight of 1. This FST is structured such that if insertions and deletions are contiguous, insertions will precede deletions \citep[Cf.][]{holmes2001evolutionary,demaio2021cumulative}.
%
(c) The COATi FST is derived via composition from the codon substitution, indel, error, and ambiguity FSTs.
}
\label{fig:coati-fst}
\end{figure}

In order to use the COATi FST to align an output sequence against an input sequence, we first convert each sequence into an acceptor, represented as a linear transducer where the input and output symbols of each transition are identical and each transition represents one nucleotide of a sequence \citep{allauzen2007openfst}. By composing the input and output acceptors with the COATi FST, we generate a transducer of all possible alignments of the two sequences. Any path through this FST represents a pairwise alignment, while the shortest path (by weight) corresponds to the best alignment. If more that one optimal alignment exists, ties are broken according to the implementation of the shortest-path algorithm. All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \citep{allauzen2007openfst}. An example of an FST-based alignment can be found in Supplementary Materials, Figure S1.

@@

We used ENSEMBL database v110 \citep{ensembl_hubbard_2002} to create an empirical dataset of protein-coding sequences for both human genes and their gorilla orthologs. We first selected human protein-coding genes that belonged to the Consensus Coding Sequence Set, that were located on an autosomal chromosome, and that had a one-to-one gorilla ortholog. We selected the canonical isoform for both species and removed any pair in which the total nucleotide length was larger larger than 6,000 nucleotides. This resulted in 14,127 sequence pairs and corresponding FASTA files containing CDS sequences. Due to the way that canonical isoforms are identified, there is no guarantee that the isoforms are orthologous even though the genes are. Therefore, a subset of the sequence pairs in this dataset contain human and gorilla sequences with different exon compositions. We have made no attempt to correct these sequence pairs because in our experience genome-wide studies rarely control for such artifacts.

@@

In order to compare COATi against other aligners, we evaluated five different alignment models: (1) COATi's FST model (i.e.\ codon-triplet-mg), (2) the amino-acid model of ClustalΩ \citep{clustal_omega_sievers_2011} via translation and reverse translation, (3) the amino-acid aware nucleotide model of MACSE \citep{ranwez_macse_2018}, (4) the DNA model of MAFFT \citep{katoh2013mafft}, and (5) the codon model of PRANK \citep{prank_loytynoja_2014}. 
COATi is not a symmetric pairwise aligner, as reference sequences are more constrained than non-reference sequences. In order to evaluate the importance of the choice for reference sequence, we also aligned sequences pairs using (6) COATi with gorilla sequences as the reference (i.e.\ COATi-rev). Together, these six methods allow us to evaluate both different alignment strategies and different software implementations. See Supplemental Materials for additional details, including results for COATi's marginal and ECM models.

@@

Alignment inference impacts the estimation of evolutionary distances. To quantify the impact of aligners on the estimation of evolutionary distances, we calculated Kimura's 2-parameter distance \citep[K2P;][]{kimura1980simple} for each sequence-pair and aligner combination.  K2P distances correct for multiple mutations at a site and takes into account differences in transition and transversion rates. K2P also assumes equal nucleotide frequencies and no variation in the rate of substitution across sites. While Kimura's 2-parameter distance is more suitable for non-coding sequences, it is straight forward to calculate and provides a quantitative measure of the evolutionary divergence between sequences. We also calculated p-distances \citep{saitou1987neighbor} for each sequence-pair and aligner combination, which are simply the proportions of sites that differ between two sequences. We used the R software package ape \citep{paradis2019ape} to calculate the distances.
%
Since aligners influence evolutionary distances based on their tendencies to insert gaps into sequences, we also quantified the phases and lengths of gaps introduced by each method as well as the fraction of nucleotides that are aligned against a match, mismatch, or gap.

@@

In order to compare COATi against other aligners on realistic datasets with known alignments, we developed a procedure to introduce realistic gap patterns into human-gorilla orthologous gene pairs that did not previous contain indels. We separated our 14,127 empirical sequence pairs into two sets: %
(1) 8261 sequence pairs that did not contain any gaps when aligned by COATi FST, ClustalΩ, MACSE, MAFFT, and PRANK, and (2) 5821 sequence pairs for which at least one of these five aligners added gaps.
(45 pairs that could not be aligned by at least one aligner were dropped.) We identified gap patterns from the sequence pairs that were aligned with gaps and randomly introduced them into the ungapped sequence pairs. We used an equal number of randomly sampled gap patterns from each aligner to produce a benchmark set of known alignments. The phases of gaps were preserved as were the spacings of any clusters of gaps. Segments of matches that were 96 nucleotides or longer were allowed to change length to accommodate the length of the ungapped sequence pair. (This criteria was recursively lowered if gap pattern did not fit the ungapped sequence pair.)

@@

As expected, COATi was not able to align 23 sequence pairs and COATi-rev was not able to align 193 sequence pairs. PRANK was not able to align 24 sequence pairs. MAFFT, MACSE, and ClustalΩ were able to align all sequence pairs, although the latter received some help from its wrapper script. Additionally, MAFFT produced alignments with columns that contained only gaps, which were removed by its wrapper script.
A total of 10,296 sequence pairs were aligned equivalently by all five methods, including 8,261 sequence pairs that did not contain any gaps, 664 that had identical alignments, and another 1,371 that had identical scores.

Compared to other aligners, COATi inferred more indels and aligned more nucleotides against a gap while producing fewer mismatches (Tab.~\ref{tab:gaps-empirical}). Additionally, COATi inferred that 60\% indels occurred in phase 1 or phase 2 in our coding sequences. MACSE's amino-acid aware nucleotide model inferred 7\%, while MAFFT's non-coding DNA model inferred 48\%. PRANK's codon and ClustalΩ's amino-acid models were constrained by their assumptions and inferred no phase 1 or phase 2 gaps. Note that ClustalΩ's gaps with phases of 1 and 2 were created by our wrapper script handling DNA sequences with lengths that are not multiples of 3.

\begin{table}[h]
\caption{\label{tab:gaps-empirical}COATi produced more indels and fewer mismatches compared to other aligners.}
\centering
\begin{tabular}[t]{lrrrccc}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Indel Phases\textsuperscript{1}} & \multicolumn{3}{c}{Homology Patterns\textsuperscript{2}}\\
\cmidrule(l{3pt}r{3pt}){2-4}
\cmidrule(l{3pt}r{3pt}){5-7}
Method & \multicolumn{1}{c}{1} & \multicolumn{1}{c}{2} & \multicolumn{1}{c}{3} & Matches & Mismatches & Gaps\\
\midrule
COATi & 4493 & 3962 & 5822 & 95.93\% & 0.79\% & 3.28\%\\
\addlinespace
ClustalΩ & 8 & 6 & 10230 & 95.93\% & 1.54\% & 2.52\%\\
\addlinespace
MACSE & 455 & 497 & 13017 & 96.13\% & 1.33\% & 2.54\%\\
\addlinespace
MAFFT & 2317 & 2631 & 5458 & 96.12\% & 1.36\% & 2.52\%\\
\addlinespace
PRANK & 0 & 0 & 10862 & 95.64\% & 0.84\% & 3.52\%\\
\addlinespace
COATi-rev & 3992 & 3775 & 6176 & 95.95\% & 0.79\% & 3.25\%\\
\bottomrule
\addlinespace
\multicolumn{7}{p{30em}}{\footnotesize\textsuperscript{1}Total number of indels inferred by each aligner separated by phase.}\\
\multicolumn{7}{p{30em}}{\footnotesize\textsuperscript{2}The percent of nucleotides that were aligned against a match, mismatch, and gap respectively. This is different than the percent of columns that contain a match, mismatch, or gap because match and mismatch columns are counted twice.}

@@

The distribution of evolutionary distances for COATi, COATi-rev, and PRANK were biologically reasonable for human-gorilla gene pairs, with means of 0.8\%--0.9\% (Fig.~\ref{fig:k2p-empirical}a). Conversely, the means for ClustalΩ, MACSE, and MAFFT were much higher, 1.5\%--1.8\%. These larger means are due to the fact that the right tails of the distributions for ClustalΩ, MACSE, and MAFFT were larger than the other methods. As expected, the alignment methods that ``overaligned'' sequences (produced more mismatches and fewer gaps) also produced higher evolutionary distances on average (Tab.~\ref{tab:gaps-empirical} and Fig.~\ref{fig:k2p-empirical}a). The differences between the distances inferred by COATi and other methods also shows ClustalΩ, MACSE, and MAFFT producing higher distance estimates for a subset of sequences (Fig.~\ref{fig:k2p-empirical}b).
%
COATi's alignments produced significantly lower K2P distances than ClustalΩ, MACSE, MAFFT, and PRANK. Single-tailed Wilcoxon signed-rank tests on matched data produced p-values $< 2.2\times10^{-16}$. Additionally, COATi's K2P distances did not depend on whether human or gorilla was used as the reference sequence (p-value of $0.4212$ for a two-sided test). Additional analysis of the empirical alignments can be found in Supplementary Methods, including results using p-distances, which did not differ from the results using K2P distances.

@@

\begin{figure}[h!]
    \centering%
    \includegraphics{figures/fig-k2p-empirical.pdf}
    \par
    \caption{COATi's alignments produce biologically reasonable evolutionary distances. (a) The distribution of K2P distances inferred from alignments generated by each method. The averages of each distribution are indicated by vertical lines. The averages are COATi = 0.0084, ClustalΩ = 0.0178, MACSE = 0.0149, MAFFT = 0.0154, PRANK = 0.0092, and COATi-rev = 0.0084. (b) The distribution of the differences between distances inferred by COATi and other methods. The X-axes of both plots have been pseudo-log transformed using the inverse hyperbolic sine.\label{fig:k2p-empirical}}
\end{figure}

@@

The vast majority of sequence pairs (12,344) were identified as showing evidence of negative selection when they were aligned using COATi, ClustalΩ, MACSE, MAFFT, PRANK, and COATi-rev (Fig.~\ref{fig:sel-empirical}). Another 1087 sequence pairs were identified as showing evidence of positive selection when they were aligned using all 6 strategies. The most common pattern that showed disagreement between aligners were 91 sequence pairs that were identified as positively selected by COATi, PRANK, and COATi-rev. The most common singleton pattern was 64 sequence pairs identified as positively selected only by MAFFT. Notably, 36 sequence pairs were identified as positively selected and 9 sequence pairs were identified as negatively selected by by COATi and COATi-rev only.
%
In total, COATi identified 1,367 (\sim{}10\%) sequence pairs as showing evidence of positive selection. ClustalΩ identified 1,228. MACSE identified 1,183. MAFFT identified 1,340. And PRANK identified 1,349. Note that we did not explore whether any of these inferences were biologically or statistically significant.

\begin{figure}[h!]
    \centering%
    \includegraphics{figures/fig-sel-empirical.pdf}
    \par
    \caption{Aligners varied in which sequence pairs they identified as undergoing positive selection.
    In this UpSet plot, the bottom panel displays the 16 most frequent intersection patterns among aligners. A black circle represents positive selection. The most frequent pattern was that no aligner found positive selection while the second most frequent pattern was that all aligners found positive selection. Other patterns involved a disagreement between aligners about whether a sequence pair showed evidence of positive or negative selection. The top panel displays the number of sequence pairs in each grouping.
%
    \label{fig:sel-empirical}}
\end{figure}

@@

COATi produced better alignments than ClustalΩ, MACSE, MAFFT, and PRANK when evaluated by our benchmark data set. It produced more best alignments, more perfect alignments, and less imperfect alignments (Tab.~\ref{table:comp}).
%
COATi was significantly more accurate (lower $d_{seq}$) at inferring the benchmark alignments compared to the other methods (Tab.~\ref{table:comp}). All p-values were significant according to the one-tailed, paired Wilcoxon signed-rank tests. (P-value of COATi vs PRANK was $5.1 \times 10^{-8}$, and p-values of COATi vs the other aligners were $< 2.2 \times 10^{-16}$). Notably, the average alignment error of the second best protocol was four times larger than COATi's. We also calculated the average distance of alignments between aligners and used principle coordinate analysis (PCoA) to project the resulting distance matrix into two dimensions (Fig.~\ref{fig:dseq-benchmarks}). COATi was the closest aligner to the benchmark, and other aligners tended to diverge from the benchmark in different directions. Because benchmark alignments were derived from gap patterns estimated by different aligners, we also performed separate PCoA analyses for each source of gap patterns (Supplementary Materials, Fig.~S6). While all aligners, except ClustalΩ, do exceptionally well on their own patterns, COATi does exceptionally well on every set of patterns.

@@

We have developed a statistical alignment program using finite-state transducers called COATi. On empirical data, COATi produced alignments that provided reasonable biological inferences. On semi-empirical benchmarks COATi was more accurate than the other aligners/alignment strategies that we tested: ClustalΩ's amino-acid model, MACSE's amino-acid aware nucleotide model, MAFFT's DNA model, and PRANK's codon model. Unlike standard codon or amino-acid alignment strategies, COATi supports aligning protein-coding sequences using all three phases of gaps. Consistent with \cite{taylor2004occurrence} and \cite{zhu2022profiling}, our COATi results indicate that only 41\% indels in protein-coding sequences between humans and gorillas are phase-3 indels. Therefore alignment strategies that only support phase-3 gaps in protein-coding sequences are sub-optimally aligning the 59\% of indels that are phase-1 and phase-2.

>>>>

%!TEX program=pdflatex
\documentclass[12pt,letterpaper]{article}
\usepackage{ifluatex}

@@

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\ifluatex
  \usepackage[warnings-off={mathtools-colon,mathtools-overbracket}]{unicode-math}
  \usepackage{fontspec}
  \setmainfont{TeX Gyre Pagella}
  \setmathfont{Latin Modern Math}
\else
  \usepackage{etex}
  \usepackage{fixltx2e}
  \usepackage{morefloats}
  \usepackage{tgpagella}
  \usepackage{textgreek}
  \usepackage[T1]{fontenc}
\fi

%\usepackage[main=english,greek]{babel}

\usepackage{microtype}

@@

% document begins here
\begin{document}
% title goes here:
\begin{flushleft}
{\Large\textbf{COATi: statistical pairwise alignment of protein-coding sequences}}
\newline
% authors go here:
\\
Juan J. Garc\'{i}a Mesa\textsuperscript{1,2},
Ziqi Zhu\textsuperscript{1,3},
Reed A. Cartwright\textsuperscript{1,3,*}\\
\bigskip
1 The Biodesign Institute, Arizona State University, Tempe, Arizona, USA\\
2 Ira A.\ Fulton Schools of Engineering, Arizona State University, Tempe, Arizona, USA\\
3 School of Life Sciences, Arizona State University, Tempe, Arizona, USA\\
\bigskip
* cartwright@asu.edu

@@

\begin{abstract}
\noindent
Sequence alignment is an essential method in bioinformatics and the basis of many analyses, including phylogenetic inference, ancestral sequence reconstruction, and gene annotation. Sequencing artifacts and errors made during genome assembly, such as abiological frameshifts and incorrect early stop codons, can impact downstream analyses leading to erroneous conclusions in comparative and functional genomic studies. More significantly, while indels can occur both within and between codons in natural sequences, most amino-acid- and codon-based aligners assume that indels only occur between codons. This mismatch between biology and alignment algorithms produces suboptimal alignments and errors in downstream analyses.

To address these issues, we present COATi, a statistical, codon-aware pairwise aligner that supports complex insertion-deletion models and can handle artifacts present in genomic data. COATi allows users to reduce the amount of discarded data while generating more accurate sequence alignments. COATi can infer indels both within and between codons, leading to improved sequence alignments. We applied COATi to a dataset containing orthologous protein-coding sequences from humans and gorillas and conclude that 41\% of indels occurred between codons, agreeing with previous work in other species. We also applied COATi to semi-empirical benchmark alignments and find that it outperforms several popular alignment programs on several measures of alignment quality and accuracy.

@@

Within protein-coding sequences, indels may occur between any pair of adjacent nucleotides, and therefore, gaps in alignments of natural sequences may occur both between and within codons (Fig.\ \ref{fig:aln}). In our nomenclature, gaps that start after the first, second, and third positions of a codon are known as phase-1, phase-2, and phase-3 gaps, respectively. Here, gaps that occur between codons are phase-3 gaps whereas they may be known as phase-0 gaps in earlier studies \cite[e.g.][]{taylor2004occurrence}. Indels that occur within codons can produce an amino-acid change in addition to adding or removing nucleotides (Fig.\ \ref{fig:aln}), and due to the structure of the genetic code, phase-1 gaps are less common than phase-2, which are less common than phase-3 \citep{taylor2004occurrence, zhu2022profiling}. While all three phases of gaps occur in natural sequences, alignments performed in amino-acid or codon space force all gaps to be phase-3 gaps. Because only about 42\% of indels are phase-3 \citep{taylor2004occurrence, zhu2022profiling}, this mismatch between aligner assumptions and biology can produce suboptimal alignments and inflated estimates of sequence divergence
\citep[Fig.\ \ref{fig:aln};][]{redelings2007incorporating}.

@@

@@

COATi FST is not a true probabilistic FST \citep{cotterell-etal-2014-stochastic} and cannot be used as-is to simulate output sequences based on an input sequence. This is because it is missing a parameter to control how often Ns are added to the output sequence. This is a design feature to allow COATi FST to properly weight ambiguous nucleotides as representing any other symbol. In addition, COATi's indel FST (Fig.~\ref{fig:coati-fst}b) implicitly assumes that match states occur both immediately before and immediately after the alignment. This allows COATi FST to assign the same weight to gaps at the beginning and end of an alignment but also introduces a normalization constant (not shown) to reflect that mass is lost by not allowing the indel FST to terminate from node C. While this normalization constant is not needed to find the most likely alignment or to sample alignments, it can be calculated using a Markov model that considers only the transitions between states M and D in the indel FST (Fig.~\ref{fig:coati-fst}b) and eliding any states in between:
\[
\frac{1-e + g(e-g)^{n+1}}{1-e+g} \approx \frac{1-e}{1-e+g} 
\]
where $g$ is the gap opening parameter, $e$ is the gap-extension parameter, and $n$ is the length of the input sequence. The exponent has a $+1$ due to the presence of an ``immortal match'' immediately after the input sequence.

@@

In order to use the COATi FST to align an output sequence against an input sequence, we first convert each sequence into an acceptor, represented as a linear transducer where the input and output symbols of each transition are identical and each transition represents one nucleotide of a sequence \citep{allauzen2007openfst}. By composing the input and output acceptors with the COATi FST, we generate a transducer of all possible alignments of the two sequences. Any path through this FST represents a pairwise alignment, while the shortest path (by weight) corresponds to the best alignment. If more than one optimal alignment exists, ties are broken according to the implementation of the shortest-path algorithm. All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \citep{allauzen2007openfst}. An example of an FST-based alignment can be found in Supplementary Materials, Figure S1.

@@

We used ENSEMBL database v110 \citep{ensembl_hubbard_2002} to create an empirical dataset of protein-coding sequences for both human genes and their gorilla orthologs. We first selected human protein-coding genes that belonged to the Consensus Coding Sequence Set, that were located on an autosomal chromosome, and that had a one-to-one gorilla ortholog. We selected the canonical isoform for both species and removed any pair in which the total nucleotide length was larger than 6,000 nucleotides. This resulted in 14,127 sequence pairs and corresponding FASTA files containing CDS sequences. Due to the way that canonical isoforms are identified, there is no guarantee that the isoforms are orthologous even though the genes are. Therefore, a subset of the sequence pairs in this dataset contain human and gorilla sequences with different exon compositions. We have made no attempt to correct these sequence pairs because in our experience genome-wide studies rarely control for such artifacts.

@@

In order to compare COATi against other aligners, we evaluated five different alignment models: (1) COATi's FST model (i.e.\ codon-triplet-mg), (2) the amino-acid model of Clustal\textOmega{}{} \citep{clustal_omega_sievers_2011} via translation and reverse translation, (3) the amino-acid aware nucleotide model of MACSE \citep{ranwez_macse_2018}, (4) the DNA model of MAFFT \citep{katoh2013mafft}, and (5) the codon model of PRANK \citep{prank_loytynoja_2014}. 
COATi is not a symmetric pairwise aligner, as reference sequences are more constrained than non-reference sequences. In order to evaluate the importance of the choice for reference sequence, we also aligned sequences pairs using (6) COATi with gorilla sequences as the reference (i.e.\ COATi-rev). Together, these six methods allow us to evaluate both different alignment strategies and different software implementations. See Supplemental Materials for additional details, including results for COATi's marginal and ECM models.

@@

Alignment inference impacts the estimation of evolutionary distances \citep{redelings2007incorporating}. To quantify the impact of aligners on the estimation of evolutionary distances, we calculated Kimura's 2-parameter distance \citep[K2P;][]{kimura1980simple} for each sequence-pair and aligner combination.  K2P distances correct for multiple mutations at a site and take into account differences in transition and transversion rates. K2P also assumes equal nucleotide frequencies and no variation in the rate of substitution across sites. While Kimura's 2-parameter distance is more suitable for non-coding sequences, it is straightforward to calculate and provides a quantitative measure of the evolutionary divergence between sequences. We also calculated p-distances \citep{saitou1987neighbor} for each sequence-pair and aligner combination, which are simply the proportions of sites that differ between two sequences. We used the R \citep{r2024} software package ape \citep{paradis2019ape} to calculate the distances.
%
Since aligners influence evolutionary distances based on their tendencies to insert gaps into sequences, we also quantified the phases and lengths of gaps introduced by each method as well as the fraction of nucleotides that are aligned against a match, mismatch, or gap.

@@

In order to compare COATi against other aligners on realistic datasets with known alignments, we developed a procedure to introduce realistic gap patterns into human-gorilla orthologous gene pairs that did not previously contain indels. We separated our 14,127 empirical sequence pairs into two sets: %
(1) 8261 sequence pairs that did not contain any gaps when aligned by COATi FST, Clustal\textOmega{}, MACSE, MAFFT, and PRANK, and (2) 5821 sequence pairs for which at least one of these five aligners added gaps.
(45 pairs that could not be aligned by at least one aligner were dropped.) We identified gap patterns from the sequence pairs that were aligned with gaps and randomly introduced them into the ungapped sequence pairs. We used an equal number of randomly sampled gap patterns from each aligner to produce a benchmark set of known alignments. The phases of gaps were preserved as were the spacings of any clusters of gaps. Segments of matches that were 96 nucleotides or longer were allowed to change length to accommodate the length of the ungapped sequence pair. (This criteria was recursively lowered if the gap pattern did not fit the ungapped sequence pair.)

@@

As expected, COATi was not able to align 23 sequence pairs and COATi-rev was not able to align 193 sequence pairs. PRANK was not able to align 24 sequence pairs. MAFFT, MACSE, and Clustal\textOmega{} were able to align all sequence pairs, although the latter received some help from its wrapper script. Additionally, MAFFT produced alignments with columns that contained only gaps, which were removed by its wrapper script.
A total of 10,296 sequence pairs were aligned equivalently by all five methods, including 8,261 sequence pairs that did not contain any gaps, 664 that had identical alignments, and another 1,371 that had identical scores.

Compared to other aligners, COATi inferred more indels and aligned more nucleotides against a gap while producing fewer mismatches (Tab.~\ref{tab:gaps-empirical}). Additionally, COATi inferred that 60\% indels occurred in phase 1 or phase 2 in our coding sequences. MACSE's amino-acid aware nucleotide model inferred 7\%, while MAFFT's non-coding DNA model inferred 48\%. PRANK's codon and Clustal\textOmega{}'s amino-acid models were constrained by their assumptions and inferred no phase 1 or phase 2 gaps. Note that Clustal\textOmega{}'s gaps with phases of 1 and 2 were created by our wrapper script handling DNA sequences with lengths that are not multiples of 3.

@@

The distribution of evolutionary distances for COATi, COATi-rev, and PRANK were biologically reasonable for human-gorilla gene pairs, with means of 0.8\%--0.9\% (Fig.~\ref{fig:k2p-empirical}a). Conversely, the means for Clustal\textOmega{}, MACSE, and MAFFT were much higher, 1.5\%--1.8\%. These larger means are due to the fact that the right tails of the distributions for Clustal\textOmega{}, MACSE, and MAFFT were larger than the other methods. As expected, the alignment methods that ``overaligned'' sequences (produced more mismatches and fewer gaps) also produced higher evolutionary distances on average (Tab.~\ref{tab:gaps-empirical} and Fig.~\ref{fig:k2p-empirical}a). The differences between the distances inferred by COATi and other methods also shows Clustal\textOmega{}, MACSE, and MAFFT producing higher distance estimates for a subset of sequences (Fig.~\ref{fig:k2p-empirical}b).
%
COATi's alignments produced significantly lower K2P distances than Clustal\textOmega{}, MACSE, MAFFT, and PRANK. Single-tailed Wilcoxon signed-rank tests on matched data produced p-values $< 2.2\times10^{-16}$. Additionally, COATi's K2P distances did not depend on whether human or gorilla was used as the reference sequence (p-value of $0.4212$ for a two-sided test). Additional analyses of the empirical alignments can be found in Supplementary Methods, including results using p-distances, which did not differ from the results using K2P distances.

@@

@@

The vast majority of sequence pairs (12,344) were identified as showing evidence of negative selection when they were aligned using COATi, Clustal\textOmega{}, MACSE, MAFFT, PRANK, and COATi-rev (Fig.~\ref{fig:sel-empirical}). Another 1087 sequence pairs were identified as showing evidence of positive selection when they were aligned using all 6 strategies. The most common pattern that showed disagreement between aligners were 91 sequence pairs that were identified as positively selected by COATi, PRANK, and COATi-rev. The most common singleton pattern was 64 sequence pairs identified as positively selected only by MAFFT. Notably, 36 sequence pairs were identified as positively selected and 9 sequence pairs were identified as negatively selected by COATi and COATi-rev only.
%
In total, COATi identified 1,367 ($\sim$10\%) sequence pairs as showing evidence of positive selection. Clustal\textOmega{} identified 1,228. MACSE identified 1,183. MAFFT identified 1,340. And PRANK identified 1,349. Note that we did not explore whether any of these inferences were biologically or statistically significant.

@@

COATi produced better alignments than Clustal\textOmega{}, MACSE, MAFFT, and PRANK when evaluated by our benchmark data set. It produced more best alignments, more perfect alignments, and less imperfect alignments (Tab.~\ref{table:comp}).
%
COATi was significantly more accurate (lower $d_{seq}$) at inferring the benchmark alignments compared to the other methods (Tab.~\ref{table:comp}). All p-values were significant according to the one-tailed, paired Wilcoxon signed-rank tests. (P-value of COATi vs PRANK was $5.1 \times 10^{-8}$, and p-values of COATi vs the other aligners were $< 2.2 \times 10^{-16}$). Notably, the average alignment error of the second best protocol was four times larger than COATi's. We also calculated the average distance of alignments between aligners and used principal coordinate analysis (PCoA) to project the resulting distance matrix into two dimensions (Fig.~\ref{fig:dseq-benchmarks}). COATi was the closest aligner to the benchmark, and other aligners tended to diverge from the benchmark in different directions. Because benchmark alignments were derived from gap patterns estimated by different aligners, we also performed separate PCoA analyses for every source of gap patterns (Supplementary Materials, Fig.~S6). While all aligners, except Clustal\textOmega{}, do exceptionally well on their own patterns, COATi does exceptionally well on every set of patterns.

@@

We have developed a statistical alignment program using finite-state transducers called COATi. On empirical data, COATi produced alignments that provided reasonable biological inferences. On semi-empirical benchmarks, COATi was more accurate than the other aligners/alignment strategies that we tested: Clustal\textOmega{}'s amino-acid model, MACSE's amino-acid aware nucleotide model, MAFFT's DNA model, and PRANK's codon model. Unlike standard codon or amino-acid alignment strategies, COATi supports aligning protein-coding sequences using all three phases of gaps. Consistent with \cite{taylor2004occurrence} and \cite{zhu2022profiling}, our COATi results indicate that only 41\% indels in protein-coding sequences are phase-3 indels between humans and gorillas. Therefore, alignment strategies that only support phase-3 gaps in protein-coding sequences are suboptimally aligning the 59\% of indels that are phase-1 and phase-2.
